Community engagement and subgroup meta-knowledge: Some factors in the soul of a community

### 2015-03-16 ResponseToReviewers.tex

\documentclass[12pt]{article}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{margin=1in}
%\geometry{ignorefoot}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage{amsmath}
\usepackage{mathrsfs}
%\usepackage[numbers,sort&compress]{natbib}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{url}
\usepackage{subfigure}

\setlength{\paperwidth}{8.5in}
\setlength{\paperheight}{11in}
\setlength{\voffset}{-0.2in}
\setlength{\topmargin}{0in}
\setlength{\headheight}{0in}
\setlength{\headsep}{0in}
\setlength{\footskip}{30pt}
\setlength{\textheight}{9.25in}
\setlength{\hoffset}{0in}
\setlength{\oddsidemargin}{0in}
\setlength{\textwidth}{6.5in}
\setlength{\parindent}{0in}
\setlength{\parskip}{9pt}

\begin{document}
{\centering
Author's Response to Reviewers' Comments \\ Submission COST-D-14-00128 \\ Amelia McNamara

}

\textbf{Author's overall response to reviewers' comments:} Thank you for giving me the opportunity to clean up my submission and resubmit. I hope that the majority of the small errors have been fixed. I have fixed many typos, expanded the captions on all plots, reformatted \verb#knitr# code to be cleaner in \LaTeX\, and clarified some of my arguments. The only comment I have not incorporated into this draft is the suggestion to change some charts to tables. This has been addressed specifically below. I feel much more solid about this submission, and hope it is satisfactory.

\textbf{Reviewers' comments:}

\textbf{The content of the paper is good, but the submission looks like it was rushed. There are a lot of very small errors in the paper that need correcting before it can be reviewed.}

\textbf{Fig 1. Maps of yearly survey percentages}
          \\  \textbf{No period.} Added.
            \\ \textbf{Explain why this is interesting or useful to know in the caption} All captions have been expanded.

 \textbf{Fig 2. Expand caption, why do we care about the number of levels on the questions} All captions have been expanded.

\textbf{Typos: behaviorsec (page 7)} Removed.

\textbf{Label text on some plots is too small to read} All plot labels have been adjusted for reading clarity.

\textbf{R output needs a little cleaning, runs outside the background box, too many decimals, but it was done in knitr which is a big thing} All knitr output has been sent through \LaTeX \,using functions like \verb#xtable()#, so output should be much cleaner.

  \textbf{Would tables be better than charts for info like fig 7, 9, 13.} These charts have been made slightly larger for easier reading. However, they have not been converted to tables. I believe it is useful to be able to graphically compare communities and groups at the same time, and I believe it is harder to do so using a table.

  \textbf{Fig 14 should have a square aspect ratio, caption doesn't match x axis label} Aspect ratio has been adjusted, and the entire section text has been clarified so captions should make more sense in the context of the axis labels.

  \textbf{Section ??? still in document} Fixed.

\end{document}

### 2015-03-16 PaperFinalDraft.Rnw
first COST commit

\documentclass[smallextended]{svjour3}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}       % onecolumn (second format)
\RequirePackage{fix-cm}
\smartqed% flush right qed marks, e.g. at end of proof
\usepackage{graphicx}
\usepackage[round]{natbib}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\journalname{Computational Statistics}
\title{Community engagement and subgroup meta-knowledge: Some factors in the soul of a community}

@@

While the Gallup Poll analysis is interesting, it does raise the question of multicollinearity, as factors that are correlated with Community Attachment may simply be correlated with one of the variables that was used to compose it, and may not actually have an impact on local GDP growth.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}

\includegraphics[width=0.99\linewidth]{figure/surveyrates-1} \hfill{}

\caption[Yearly survey percentages, displayed on a log scale]{Yearly survey percentages, displayed on a log scale. Notice that some communities are always over-surveyed (for example, Palm Beach, FL) and some always appear under-surveyed (for example, Long Beach, CA). Population data compiled by the Census Bureau for Intercensal population estimates}\label{fig:surveyrates}
\end{figure}

\end{knitrout}
As mentioned above, the data were collected by Gallup through telephone surveys in 2008, 2009, and 2010. Participants were a random sample of adults living in 26 ``communities" (cities or metro areas of the United States). The data from 2008 and 2009 had 13822 and 13728 responses, respectively, while the data from 2010 contained 20271 observations. Because the data sets surveyed the same 26 communities, we can calculate the average number of survey participants in each community. In 2008, that average was 531 people, in 2009, 528 people and in 2010, 779 people. The difference in average number of survey participants will be discussed further in Section \ref{missingdata}.

@@

The varied lengths of response scales and the different phrasing of scales even with the same length suggests that this survey was quite long and complex to complete. And though the survey maintains the scale lengths for individual questions over the years, Gallup rescales all the questions down to a 3-point scale to make their analysis simpler. The complete data set provided by the Knight Foundation includes between 156 and 206 variables, depending on the year, but fully half of them are rescaled versions of the original questions. While some researchers have suggested that a 3-point scale is enough \citep{JacMat1971}, discarding data seems wasteful, especially if participants have gone to the trouble of rating on a 5- or 7-point scale. So, the remainder of this analysis works on the unscaled variables. 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}

{\centering \includegraphics[width=0.99\linewidth]{figure/LikertSizes-1}

}

\caption[Scale length distributions for each year of the survey]{Scale length distributions for each year of the survey. Notice that the distribution from 2008 is centered around 7, and the 2009 and 2010 distributions are centered around 5. All three distributions have large variation, suggesting that the Knight surveys were quite complex.}\label{fig:LikertSizes}
\end{figure}

\end{knitrout}

@@

In order to explore this, we study the pattern of missing data within the 2010 data set. This data set includes a large number of responses that are extremely sparse, containing almost no completed survey questions, but generally completed demographic information. For example, the question ``How likely are you to recommend this community to a friend or associate as a place to live?'' has 4947 missing values, and the question ``The overall quality of the colleges and universities'' has 5200 missing values. Just between those two questions, the overlap of missing values for both questions is 4887. In other words, almost all the entries that are missing a response to the question about recommending the community are also missing a response to the question about colleges. Similarly, there are 4887 respondents missing both the question about colleges and the question ``Are you registered to vote?''

@@

% latex table generated in R 3.1.2 by xtable 1.7-3 package
% Fri Mar 13 20:54:26 2015
\begin{table}[ht]
\centering
\begin{tabular}{rrrr}
  \hline
Year & Missing & Total & Percent \\ 
  \hline
2008 &   6 & 13822 & 0.04 \\ 
  2009 &  63 & 13728 & 0.46 \\ 
  2010 & 4947 & 20271 & 24.40 \\ 
   \hline
\end{tabular}
\caption{Percent of missing data from surveys. The 2010 survey has almost 25 percent missing data. The total number of entries in 2010 is also much larger, suggesting that the 2008 and 2009 datasets used a different criteria for inclusion of entries.} 
\label{missingtable}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}

{\centering \includegraphics[width=0.99\linewidth]{figure/communitysatplot-1}

}

\caption[Responses to the question, ``Taking everything into account, how satisified are you with this community as a place to live?" Communities are ordered by percentage of positive responses in 2008, making it clear the differences in distribution in 2009 and 2010]{Responses to the question, ``Taking everything into account, how satisified are you with this community as a place to live?" Communities are ordered by percentage of positive responses in 2008, making it clear the differences in distribution in 2009 and 2010.}\label{fig:communitysatplot}
\end{figure}

\end{knitrout}

Overall, the total percentage of positive responses in 2008 ranges between 58\% on the low end, and 90\% at the high end. These numbers are the overall percentage of people in each community that are answering the question, ``Taking everything into account, how satisfied are you with this community as a place to live?'' positively (this includes half of the responses labeled 3).

Communities are ordered by overall positive responses in 2008, which allows for comparisons between communities and across years. The overall trend is fairly stable, but there is some variation, both positive and negative. Looking at Figure \ref{fig:communitysatplot}, we can see that people in State College, PA typically report much greater levels of community satisfaction than people in Detroit, MI or Gary, IN. After 2008, Macon, GA sees a decrease in overall community satisfaction, moving from 73\% in 2008 to 61\% in 2009 and slightly up to 63\% in 2010. Over the same time period, Bradenton, FL shows a slight increase in community satisfaction, from 84\% in 2008 to about  88\% in 2009 and 2010.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}

{\centering \includegraphics[width=0.99\linewidth]{figure/YNall-1}

}

\caption[Responses to yes/no questions about participants' behaviors, comparing all three survey years]{Responses to yes/no questions about participants' behaviors, comparing all three survey years.}\label{fig:YNall}
\end{figure}

\end{knitrout}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}

{\centering \includegraphics[width=0.99\linewidth]{figure/YN2010plot-1}

}

\caption{Percentage difference from overall survey rates (2010 data). The letters A-J represent the activities listed in Figure \ref{fig:YNall}. A: Registered to vote, B: Voted in a local election, C: Donated money to a local organization, D: Attended a local event, E: Gave money or food to an individual, F: Participated in a church event, G: Performed local volunteer work, H: Worked with other residents to make change, I: Attended a local public meeting, J: Provided free shelter to an individual.}\label{fig:YN2010plot}
\end{figure}

\end{knitrout}

>>>>

\documentclass[smallextended]{svjour3}       % onecolumn (second format)
\RequirePackage{fix-cm}
\smartqed% flush right qed marks, e.g. at end of proof
\usepackage{graphicx}
\usepackage[round]{natbib}

\begin{document}
\journalname{Computational Statistics}
\title{Community engagement and subgroup meta-knowledge: Some factors in the soul of a community}

@@

While the Gallup Poll analysis is interesting, it does raise the question of multicollinearity, as factors that are correlated with Community Attachment may simply be correlated with one of the variables that was used to compose it, and may not actually have an impact on local GDP growth. 
<<packages, echo=FALSE>>=
require(knitr)
@ 
<<chunkoptions, echo=FALSE, include=FALSE>>=
opts_chunk$set(echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align='center')
require(dplyr)
require(ggplot2)
require(scales)
require(RColorBrewer)
require(ggmap)
require(xtable)
require(reshape2)
@
<<readdata>>=
sotc10 <- read.csv("data/sotc10.csv")
sotc09 <- read.csv("data/sotc09.csv")
sotc08 <- read.csv("data/sotc08.csv")
@
<<renaming>>=
renameData <- function(datasetname){
  plyr::rename(datasetname,c("INTRO1" = "gender", "QSB" = "citystate", "QS3"="county", "QS3_02"="county2", "QS3A"="zipcode", "QD1"="age", "QS4"="area_descrip", "QS5"="further", "QS5_2"="further2", "QN1A"="current_life", "QN1B"="future_life", "QCE1"="community_sat","QCE2"="recommend","Q3A"="proud", "Q3B"="perfect", "Q3C"="reputation", "Q4_1"="problem_1", "Q4_2"="problem_2", "Q4_3"="problem_3", "Q4_1_1"="problem_1", "Q4_1_2"="problem_2", "Q4_1_3"="problem_3","Q4_2_1"="problem_4", "Q4_2_2"="problem_5", "Q4_2_3"="problem_6","Q4_3_1"="problem_7", "Q4_3_2"="problem_8", "Q4_3_3"="problem_9","Q5"="rather_live", "Q6"= "five_years_ago", "Q6A" = "five_years_future", "Q7A"= "parks", "Q7B"= "beauty", "Q7C"= "highways", "Q7D"= "affordable_housing", "Q7E" =  "job_opportunites", "Q7F"=  "public_schools", "Q7G" = "colleges", "Q7H"= "nightlife", "Q7I"= "friends", "Q7J" = "raise_kids", "Q7K" = "heathcare", "Q7L" ="leadership", "Q7M" = "care", "Q7N"= "police", "Q7O" = "arts", "Q7P" =  "community_events", "Q8A" = "talented_grads", "Q8B" = "immigrants", "Q8C"= "minorities", "Q8D"= "families_kids", "Q8E" = "gay_lesbian", "Q8F" = "seniors", "Q8G" = "young_adults", "Q9" = "economy", "Q10" = "econ_better", "Q10A" = "trust_government", "Q11" = "employment", "Q12" = "commute", "Q13" = "job_satisfaction", "Q14" = "company_workforce", "Q15" = "enough_income", "Q15AA" = "area_hiring", "Q15AB" = "representative_leaders", "Q16A" = "treated_respectfully", "Q16B" = "well_rested", "Q16C" = "high_stress", "Q16D" = "learned_yest", "Q17" = "stress_source", "Q18" = "night_safety", "Q19" = "crime", "Q20" = "crime_increasing", "Q21"=  "registered_vote", "Q22A" = "volunteer", "Q22B" = "public_meeting", "Q22C"= "voted_local", "Q22D" = "worked_change", "Q22E" = "church", "Q22F" = "festival", "Q22G" = "donated_org", "Q22H" = "donated_individual", "Q22I" = "donated_shelter", "Q22_A" = "impact", "Q23" = "clubs", "Q24" = "friends_nearby", "Q24A" =  "friends_friends", "Q25"= "family_nearby", "Q26"= "talk_neighbors", "QD2" = "years_lived", "QD2A" ="permanent_resident", "QD3" = "work_cat", "QD3A" = "household_adults", "QD4" = "children", "QD5A" = "under6", "QD5B" = "six_12", "QD5C" = "thirteen_17", "QD6" = "married", "QD7"= "education", "QD8" = "own_home", "QD9"= "income", "QD10" = "hispanic", "QD111" = "race", "QD112" = "race2", "QD113" = "race3", "QD12" = "white_hispanic", "QD13" = "phone_numbers", "QD13A" = "cell_only", "QD13B" = "have_cell", "QD13C" = "cell_calls"))
}

sotc10 <- renameData(sotc10)
sotc09 <- renameData(sotc09)
sotc08 <- renameData(sotc08)
@

<<usedalot>>=
n1 <- c("citystate", "InGroup", "Response", "n", "Freq", "year")
colors2 <- c("#f1a340", "#998ec3")
colors1 <- brewer.pal(5, "Spectral")
colorsA <- colors1[c(2,3,4,1,5)]
colorsB <- colors1[c(2,3,4,5,1)]
@

<<LikertDataFunction>>=
LikertData <- function(positive, negative, neutral=NULL, dataset){
  pos <- dataset[dataset$Response %in% positive,]
  neg <- dataset[dataset$Response %in% negative,]
  if (is.null(neutral)==FALSE){
    neu <- dataset[dataset$Response %in% neutral, ]
    neu$Freq <- neu$Freq/2
    pos <- rbind(pos, neu)
    neg <- rbind(neg, neu)
  } else {
  neu <- NULL
  }
  neg$Freq <- -neg$Freq
  neg$Response <- factor(neg$Response)
  neg$Response <- factor(neg$Response, levels=rev(levels(neg$Response)))
  pos$Response <- factor(pos$Response)
  return(list(pos=pos, neu=neu, neg=neg))
}
@

<<calculatingInGroupFeelings>>=
FeelingsC <- function(dataset){
  Feelings <- dataset %>% group_by(citystate, InGroup) %>% summarise(percents= sum(Freq))
  Feelings <- data.frame(Feelings)
  newDF <- data.frame(InGroup=NA, OutGroup=NA, citystate=levels(Feelings$citystate))
  newDF$ratio <- (Feelings[Feelings$InGroup=="FALSE" | Feelings$InGroup=="No",]$percents) - (Feelings[Feelings$InGroup=="TRUE" | Feelings$InGroup=="Yes",]$percents)
  newDF$InGroup <- Feelings[Feelings$InGroup=="TRUE" | Feelings$InGroup=="Yes",]$percents
  newDF$OutGroup <- Feelings[Feelings$InGroup=="FALSE" | Feelings$InGroup=="No",]$percents
  return(newDF)
}
@

@@

<<gettinggeodata>>=
placesPop <- dget("data/popdata.robj")

percent1 <- summary(sotc10$citystate)/placesPop$CENSUS2010POP
percent1 <-c(percent1, summary(sotc09$citystate)/placesPop$POPESTIMATE2009)
percent1 <- c(percent1, summary(sotc08$citystate)/placesPop$POPESTIMATE2008)

popData <- data.frame(year=c(rep(2010, 26), rep(2009, 26), rep(2008, 26)), percent=percent1)
popData$percent <- popData$percent*100
popData$citystate <- c(rep(levels(sotc10$citystate),3))
@

<<surveyrates, fig.align="left", fig.cap="Yearly survey percentages, displayed on a log scale. Notice that some communities are always over-surveyed (for example, Palm Beach, FL) and some always appear under-surveyed (for example, Long Beach, CA). Population data compiled by the Census Bureau for Intercensal population estimates", out.width="0.99\\linewidth",fig.width = 9.5,fig.height=7.5>>=

p <- ggplot(aes(x=year, y=percent), data=popData)+facet_wrap(~citystate, ncol=5)+geom_line()+scale_x_continuous(breaks=c(2008, 2008.5, 2009, 2009.5, 2010), labels=c("2008","", "2009", "", "2010"))+scale_y_log10(breaks=c(0.04, 0.4, 4))+xlab("")+ylab("Percent of community surveyed")
p
@
As mentioned above, the data were collected by Gallup through telephone surveys in 2008, 2009, and 2010. Participants were a random sample of adults living in 26 ``communities" (cities or metro areas of the United States). The data from 2008 and 2009 had \Sexpr{dim(sotc08)[1]} and \Sexpr{dim(sotc09)[1]} responses, respectively, while the data from 2010 contained \Sexpr{dim(sotc10)[1]} observations. Because the data sets surveyed the same 26 communities, we can calculate the average number of survey participants in each community. In 2008, that average was \Sexpr{floor(dim(sotc08)[1]/26)} people, in 2009, \Sexpr{floor(dim(sotc09)[1]/26)} people and in 2010, \Sexpr{floor(dim(sotc10)[1]/26)} people. The difference in average number of survey participants will be discussed further in Section \ref{missingdata}.

@@

<<numlevels>>=
pullLevels <- function(dataset, range){
  numLevels <- NA
  for (i in range){
    numLevels[i] <- length(levels(as.factor(dataset[,i])))
  }
  return(numLevels)
}

#Removing the ones that don't make sense-- listing problems, in particular

# For sotc10, this works:
numLevels10 <- pullLevels(sotc10, c(11:80))
numLevels10[which(numLevels10>25)] <- NA

#For sotc09
numLevels09 <- pullLevels(sotc09, c(11:70))
numLevels09[16:24] <- NA

#For sotc08
numLevels08 <- pullLevels(sotc08, c(9:64))
# This takes out the problem questions as well as the commute question
numLevels08[which(numLevels08>20)] <- NA

numLevels <- data.frame(levels=c(numLevels08, numLevels09, numLevels10), year=c(rep(2008, times=length(numLevels08)), rep(2009, times=length(numLevels09)), rep(2010, times=length(numLevels10))))
@

The varied lengths of response scales and the different phrasing of scales even with the same length suggests that this survey was quite long and complex to complete. And though the survey maintains the scale lengths for individual questions over the years, Gallup rescales all the questions down to a 3-point scale to make their analysis simpler. The complete data set provided by the Knight Foundation includes between \Sexpr{dim(sotc08)[2]} and \Sexpr{dim(sotc10)[2]} variables, depending on the year, but fully half of them are rescaled versions of the original questions. While some researchers have suggested that a 3-point scale is enough \citep{JacMat1971}, discarding data seems wasteful, especially if participants have gone to the trouble of rating on a 5- or 7-point scale. So, the remainder of this analysis works on the unscaled variables. 
<<LikertSizes, fig.cap="Scale length distributions for each year of the survey. Notice that the distribution from 2008 is centered around 7, and the 2009 and 2010 distributions are centered around 5. All three distributions have large variation, suggesting that the Knight surveys were quite complex.",out.width="0.99\\linewidth",fig.width = 9.5,fig.height=4.5>>=
resp<- ggplot(numLevels, aes(x=levels))+geom_histogram(breaks=seq(from=0, to=15, by=1))+facet_grid(.~year)+labs(x="Number of levels on the scale", y="Number of questions using the scale")+theme(strip.text=element_text(size=14))
#title="Distribution of scale lengths used in surveys"
resp
@

@@

<<missingmatch>>=
matchup1 <- length(which(which(is.na(sotc10$recommend)) %in% which(is.na(sotc10$college))))
matchup2 <- length(which(which(is.na(sotc10$college)) %in% which(is.na(sotc10$registered_vote))))
@

In order to explore this, we study the pattern of missing data within the 2010 data set. This data set includes a large number of responses that are extremely sparse, containing almost no completed survey questions, but generally completed demographic information. For example, the question ``How likely are you to recommend this community to a friend or associate as a place to live?'' has \Sexpr{sum(is.na(sotc10$recommend)==TRUE)} missing values, and the question ``The overall quality of the colleges and universities'' has \Sexpr{sum(is.na(sotc10$college)==TRUE)} missing values. Just between those two questions, the overlap of missing values for both questions is \Sexpr{matchup1}. In other words, almost all the entries that are missing a response to the question about recommending the community are also missing a response to the question about colleges. Similarly, there are \Sexpr{matchup2} respondents missing both the question about colleges and the question ``Are you registered to vote?''

@@

<<missingdatatable, results="asis">>=
missing10 <- group_by(sotc10, recommend == "NA") %>% summarise(n = n()) 
missing09 <- group_by(sotc09, recommend == "NA") %>% summarise(n = n()) 
missing08 <- group_by(sotc08, recommend == "(Refused)") %>% summarise(n = n()) 
missing <- c(missing08$n[2], missing09$n[2], missing10$n[2])
missingtable <- as.matrix(data.frame(Year=c(2008, 2009, 2010), Missing=missing, Total=c(dim(sotc08)[1], dim(sotc09)[1], dim(sotc10)[1]), Percent=missing/c(dim(sotc08)[1], dim(sotc09)[1], dim(sotc10)[1])*100))
rownames(missingtable) <- NULL
print(xtable(x=missingtable,caption="Percent of missing data from surveys. The 2010 survey has almost 25 percent missing data. The total number of entries in 2010 is also much larger, suggesting that the 2008 and 2009 datasets used a different criteria for inclusion of entries.", display=c("d", "d", "d","d", "f"), label="missingtable"),include.rownames=FALSE)
@

@@

<<communitysat>>=
tmp1 <- tbl_df(sotc08) %>% group_by(citystate, community_sat) %>% summarise(n = n()) %>% mutate(frac=prop.table(n), year=2008)
tmp2 <- tbl_df(sotc09) %>% group_by(citystate, community_sat) %>% summarise(n = n()) %>% mutate(frac=prop.table(n), year=2009)
tmp3 <- tbl_df(sotc10) %>% group_by(citystate, community_sat) %>% summarise(n = n()) %>% mutate(frac=prop.table(n), year=2010)

tmp <- rbind(tmp1, tmp2, tmp3)
names(tmp) <- c("citystate", "Response", "n", "Freq", "year")

#Now all I have to do is make the plot
neutral <- c("3")
positive <- c("Extremely satisfied", "4")
negative <- c("2", "Not at all satisfied")

trial2 <- LikertData(positive, negative, neutral, tmp)

extremelysat08 <- tbl_df(trial2$pos) %>% filter(year==2008) %>% group_by(citystate) %>% summarise(tot=sum(Freq)) 
extremelysat <- tbl_df(trial2$pos) %>% group_by(citystate, year) %>% summarise(tot=sum(Freq)) 
mostsat <- order(extremelysat08$tot)

masterData <- data.frame(citystate=extremelysat$citystate, year=extremelysat$year, communitysat=extremelysat$tot)

trial2$pos$citystate <- factor(trial2$pos$citystate, levels=levels(trial2$pos$citystate)[mostsat])
trial2$neg$citystate <- factor(trial2$neg$citystate, levels=levels(trial2$neg$citystate)[mostsat])
trial2$neg$Response <- factor(trial2$neg$Response)
trial2$pos$Response <- factor(trial2$pos$Response)
trial2$neg$Response <- factor(trial2$neg$Response, levels=levels(trial2$neg$Response)[c(2,3,1)])

posform <- trial2$pos

maconsat <- tbl_df(posform) %>% filter(citystate=="Macon, GA") %>% group_by(year) %>% summarise(totalper=sum(Freq))
brandetonsat <- tbl_df(posform) %>% filter(citystate=="Bradenton, FL") %>% group_by(year) %>% summarise(totalper=sum(Freq))
@

<<communitysatplot, fig.cap="Responses to the question, \`\`Taking everything into account, how satisified are you with this community as a place to live?\" Communities are ordered by percentage of positive responses in 2008, making it clear the differences in distribution in 2009 and 2010.", out.width="0.99\\linewidth",fig.width = 9.5,fig.height=15,>>=
baseplot <- ggplot() + aes(x=citystate, y=Freq, fill = Response, order=Response)+facet_wrap(~year, nrow=3)
baseplot <- baseplot  +geom_bar(data = trial2$neg, stat = "identity") + scale_fill_manual(breaks=c("Not at all satisfied", "2", "3", "4", "Extremely satisfied"), values=colorsB, name="Response") + geom_bar(data = trial2$pos, stat = "identity")
baseplot <- baseplot +coord_flip() + ggtitle("Community satisfaction") + xlab("") +ylab("")
#+ggtitle("Community satisfaction")
baseplot <- baseplot + scale_y_continuous(limits=c(-0.5, 1), breaks=seq(from=-0.5, to=0.75, by=0.25), labels=c("50%", "25%", "0", "25%", "50%", "75%"))+theme(legend.text=element_text(size=14), legend.title=element_text(size=16), axis.text=element_text(size=14), strip.text=element_text(size=14))
baseplot 
@

Overall, the total percentage of positive responses in 2008 ranges between \Sexpr{round(extremelysat08[extremelysat08$citystate=="Gary, IN",]$tot*100)}\% on the low end, and \Sexpr{round(extremelysat08[extremelysat08$citystate=="State College, PA",]$tot*100)}\% at the high end. These numbers are the overall percentage of people in each community that are answering the question, ``Taking everything into account, how satisfied are you with this community as a place to live?'' positively (this includes half of the responses labeled 3).

Communities are ordered by overall positive responses in 2008, which allows for comparisons between communities and across years. The overall trend is fairly stable, but there is some variation, both positive and negative. Looking at Figure \ref{fig:communitysatplot}, we can see that people in State College, PA typically report much greater levels of community satisfaction than people in Detroit, MI or Gary, IN. After 2008, Macon, GA sees a decrease in overall community satisfaction, moving from \Sexpr{round(maconsat$totalper[1]*100)}\% in 2008 to \Sexpr{round(maconsat$totalper[2]*100)}\% in 2009 and slightly up to \Sexpr{round(maconsat$totalper[3]*100)}\% in 2010. Over the same time period, Bradenton, FL shows a slight increase in community satisfaction, from \Sexpr{round(brandetonsat$totalper[1]*100)}\% in 2008 to about  \Sexpr{round(brandetonsat$totalper[2]*100)}\% in 2009 and 2010.

@@

<<twolevel>>=
twolevel <- c("registered_vote", "volunteer", "public_meeting", "voted_local", "worked_change", "church", "festival", "donated_org", "donated_individual", "donated_shelter")

# Function that takes a dataset and calculates the percentages of cases overall
findPercen <- function(dataset){
  dfall <- data.frame()
  for (i in 1:length(twolevel)){
  if(twolevel[i] %in% names(dataset)){
      tmp <- data.frame(Question=twolevel[i], prop.table(table(dataset[,twolevel[i]])))
      dfall <- rbind(dfall, tmp)
  }
  else{
    tmp <- data.frame(Question=twolevel[i], Var1 = c("No", "Yes"), Freq=c(NA, NA))
    dfall <- rbind(dfall, tmp)
  }
}
names(dfall) <- c("Question", "Response", "Freq")
return(dfall)
}
#dplyr would be much better here, but can't quite figure out how to forumulate it.

df10 <- findPercen(sotc10)
df10$Year <- 2010
df09 <- findPercen(sotc09)
df09$Year <- 2009
df08 <- findPercen(sotc08)
df08$Year <- 2008

dfall <- rbind(df10, df09, df08)
dfall$Question <- factor(dfall$Question, levels=levels(dfall$Question)[order(df10$Freq[df10$Response=="Yes"])])
positive <- "Yes"
negative <- "No"
trial1 <- LikertData(positive, negative, neutral=NULL, dataset=dfall) 
@

<<YNall, out.width="0.99\\linewidth",fig.width = 15,fig.height=9.5,fig.cap="Responses to yes/no questions about participants' behaviors, comparing all three survey years.">>=
baseplot <- ggplot() + aes(Question, Freq, fill = Response) +facet_wrap(~Year)
baseplot <- baseplot + geom_bar(data = trial1$neg, stat = "identity") + geom_bar(data = trial1$pos, stat = "identity")
baseplot <- baseplot + scale_y_continuous(breaks=seq(from=-0.75, to=0.75, by=0.25), labels=percent)
#+ggtitle("Yes/no questions") 
baseplot <- baseplot + coord_flip()+ xlab("")
baseplot <- baseplot + scale_x_discrete(labels=rev(c("Registered to vote \n", "Voted in local election \n", "Donated money to a \n local organization", "Attended a local \n festival or event", "Gave money or food \n to an individual", "Participated in a \n church event", "Performed local \n volunteer work", "Worked with other residents \n to make change", "Attended a local \n public meeting", "Provided free shelter \n to an individual")))
baseplot <- baseplot + ylab("")+theme(legend.text=element_text(size=14), legend.title=element_text(size=16), axis.text.y=element_text(size=20), title=element_text(size=24), axis.text.x=element_text(size=14), strip.text=element_text(size=18))
baseplot
@

<<YN2010>>=
dfall <- data.frame()
for (i in 1:length(twolevel)){
  tmp <- data.frame(Question=twolevel[i], prop.table(table(sotc10[,twolevel[i]])))
  dfall <- rbind(dfall, tmp)
}
names(dfall) <- c("Question", "Response", "Freq")

citystateResp <- function(dataset){
  # Create a loop that calculates the percentages of cases in each factor level for each city
  dfcities <- data.frame()
  questionsIn <- which(twolevel %in% names(dataset))
  for (j in 1:length(levels(dataset$citystate))){
    thiscity <- subset(dataset, citystate==levels(dataset$citystate)[j])
    for (i in questionsIn){
      tmp <- data.frame(levels(dataset$citystate)[j], Question=twolevel[i], prop.table(table(thiscity[,twolevel[i]])))
      dfcities <- rbind(dfcities, tmp)
    }
  }
  names(dfcities) <- c("citystate", "Question", "Response", "Freq")

  # Now we calculate difference between the overall and the cities to see where the variation comes from
  dfcities$diff <- 0
  for (i in 1:length(levels(dfcities$Question))){
    # Yes
    overallrate <- dfall[dfall$Question==levels(dfall$Question)[i] & dfall$Response=="Yes",]$Freq
    dfcities$diff[(dfcities$Question==levels(dfcities$Question)[i]& dfcities$Response=="Yes")] <- (overallrate - dfcities[(dfcities$Question==levels(dfcities$Question)[i] & dfcities$Response=="Yes"),]$Freq)/overallrate
  }
  return(dfcities)
}

dfcities10 <- citystateResp(sotc10)
dfcitiesYes <- dfcities10[dfcities10$Response=="Yes",]
dfcitiesYes$col <- dfcitiesYes$dif>0
dfcitiesYes$Question <- factor(dfcitiesYes$Question, levels=levels(dfcitiesYes$Question)[order(dfall[dfall$Response=="Yes",]$Freq)])
@

<<startingoverall>>=
dfcities09 <- citystateResp(sotc09) %>% mutate(year="2009")
dfcities08 <- citystateResp(sotc08) %>% mutate(year="2008")
dfcities10 <- dfcities10 %>% mutate(year="2010")
md1 <-rbind(dfcities10, dfcities09, dfcities08) %>% filter(Response=="Yes") %>% select(-Response, -diff) %>% filter(Question %in% c("registered_vote", "voted_local", "volunteer", "worked_change", "public_meeting"))
md1 <- melt(md1)
md1 <- dcast(md1, citystate + year ~ Question, mean)
md1$year <- as.numeric(md1$year)
masterData <- left_join(masterData, md1)
@
